---
title: "Exercise: Continuous Random Variables"
format:
  html:
    embed-resources: true

---

We have considered the two-person meeting problem, but what if there are more people? 
Suppose that *three* people meet for lunch between noon and 1:00. 
Measure their arrival times in fractions of an hour after noon, e.g. 0.25 represents 15 minutes after noon, so that each arrival time takes values in the continuous interval [0, 1]. 
Assume that the three arrival times follow a Uniform(0, 1) distribution, independently. 
Let $X$ be the time (fractions of an hour after noon) at which the first person arrives. 
It can be shown that the cdf of $X$ is
$$
F(x) = 1 - (1-x)^3, \qquad 0<x<1
$$
(You'll see some justification for this fact by the end of this exercise, and further justification in an upcoming exercise.)

1.  Evaluate and interpret in context $F(0.5)$.
1.  Evaluate and interpret in context $F(0.5) - F(0.2)$
1.  Sketch a histogram for $X$ with 10 bins of equal width.
1.  Use the histogram to sketch the general shape of the probability density function (pdf) of $X$.
1.  Find and interpret in context the 25th percentile of $X$.
1.  Find the quantile function of $X$.
1.  Sketch a spinner for simulating values of $X$. 
Label the values at "3, 6, 9 o'clock", and indicate where the value 0.5 should be placed.
1.  Conduct a simulation to approximate the distribution of $X$ and the quantities from the previous parts. 
Based on the simulation results, does $F$ seem like the appropriate cdf?

```{r}
#| echo: false
#| eval: true

knitr::knit_exit()
```

{{< include _r_setup.qmd >}}

{{< include _python_setup.qmd >}}

## Solutions

```{python}
P = Uniform(0, 1) ** 3

X = RV(P).apply(min)

x = X.sim(10000)

x.plot()

Beta(1, 3).plot(); # this is the theoretical distribution
```


```{python}
x.count_lt(0.5) / 10000
```

```{python}
x.count_lt(0.2) / 10000
```

```{python}
x.plot(bins = 10)
```

```{python}
x.quantile(0.25)
```

```{python}
x.quantile(0.75)
```




```{r}
#| label: fig-meeting-nonuniform-spinner
#| echo: false
#| fig-cap: "The same spinner is displayed on both sides, with different features highlighted on the left and right. Only selected rounded values are displayed, but in the idealized model the spinner is infinitely precise so that any real number is a possible outcome. Notice that the values on the axis are *not* evenly spaced."
#| fig-subcap: 
#|   - "Values at 1 o'clock, 2 o'clock, 3 o'clock, etc, corresponding to the 8.33th, 16.67th, 25th, etc, percentiles"
#|   - "Spinner axis marked in increments of 0.1 from 0 to 1"
#| layout-ncol: 2

n = 12

xp <- data.frame(
  x = (0:(n-1))/n,
  p = rep(1/n, n)
  )

cdf = c(0, cumsum(xp$p))

plotp = (cdf[-1] + cdf[-length(cdf)]) / 2
  
spinner1 <- ggplot(xp, aes(x="", y=p, fill=x))+
  geom_bar(width = 1, stat = "identity", color="black", fill="white") + 
  coord_polar("y", start=0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = xp$x, minor_breaks = (0:99)/100, labels=c("1|0  ", round(qbeta((1:(n-1))/n, shape1 = 1, shape2 = 3), 3))) +
  theme(axis.text.x=element_text(size=12, face="bold"))

spinner1


xr = seq(0.1, 1, 0.1)
xl = xr - 0.1

p = (1 - xl) ^ 3 - (1 - xr) ^ 3

xp <- data.frame(xr, p)

cdf = c(0, cumsum(xp$p))



plotp = (cdf[-1] + cdf[-length(cdf)]) / 2
  
spinner2 <- ggplot(xp, aes(x="", y=p, fill=xr))+
  geom_bar(width = 1, stat = "identity", color="black", fill="white", linetype=1) + 
  coord_polar("y", start=0) +
  # coord_curvedpolar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = cdf[-c(length(cdf))], labels=c("1|0  ", seq(0.1, 0.9, 0.1))) +
  theme(axis.text.x=element_text(angle = 0, size=12, face = "bold")) +
    geom_text(aes(y = plotp,
                label = percent(p, accuracy = 0.1)), size=4, color=c(rep("black",6), rep(NA, 4)))

spinner2

```
